A simulation model of DNA methylation data

### 2015-05-06 Rmarkdown/methsim.Rmd
Edits from final proofread. How do I keep finding freaking typos!?

There is a lack of gold-standard datasets in the field of DNA methylation for the benchmarking and validation of analysis methods. Therefore, the development of realistic, cheap and user-friendly software to simulate DNA methylation data will be of benefit.

@@

`Sherman` ([http://www.bioinformatics.babraham.ac.uk/projects/sherman/](http://www.bioinformatics.babraham.ac.uk/projects/sherman/)) is software to simulate bisulfite-sequencing reads as `FASTQ` files, including various 'contaminants', such as SNPs, basecall errors and sequence artefacts. The simulated reads are designed for comparing the performance of different alignment strategies. `Sherman` has many parameters, of which the ones relevant to our discussion of simulating realistic DNA methylation data are `-CG` and `-CH`, the bisulfite conversion rates for CG and CH methylation loci, respectively. These are set by the user with values between $0$ and $100$ ($\%$). Reads are simulated by sampling from the user-specified reference genome. When a read contains a CG (resp. CH) locus, it is randomly assigned as being converted (unmethylated) with probability `-CG` / 100 (resp. `-CH` / 100).

@@

The most widely studied downstream analysis problem is that of identifying differential methylation, which is done by comparing summary measures of methylation, such as $\beta$-values, between two or more groups. It is therefore generally not necessary to simulate individual reads, but is sufficient to directly simulate these aggregated measures, when simulating data for these studies.

Most papers that propose a new method for downstream analysis of bisulfite-sequencing data include a simulation section. Generally, such a simulation method exists to support claims about the performance of the proposed method and is not a major feature of the paper. Consequently, the simulation model is often only briefly described and a software implementation is rarely made available. In fact, until recently, of the methods reviewed in this section, only that of \citet{Lacey:2013iy} had a software implementation available. As I was writing this chapter, the `WGBSSuite` software was published \citep{Rackham:2015bv}. `WGBSSuite` is the only software specifically published for the purpose of simulating whole-genome bisulfite-sequencing data for comparing methods for identifying differential methylation. `WGBSSuite` is available for download as a collection of R scripts[^WGBSSuite].

@@

Almost all downstream analyses are _reference-based_ (see Section \ref{sec:methylated_or_unmethylated}), so it is desirable to know how these methods perform with respect to the relevant reference genome, not a simulated genome in which the location of methylation loci vary from simulation to simulation.

@@

Recall that $B_{i, j}$ is the underlying _true_ methylation level at locus $i$ in sample $j$. In the context of simulating data for comparing differential methylation calling methods, we want these to be group-specific. That is, we want to specify $B_{i, j_{k}}$, where $j_{k}$ indicates that sample $j$ comes from group $k$. For non-differentially methylated loci all $k$ groups have identical $B_{i, j_{k}} = B_{i, j_{0}}$; differentially methylated loci are simulated by setting $B_{i, j_{k}} \neq B_{i, j_{k'}}$ for some $k \neq k'$.

Under the beta-binomial model, the true methylation level, $B_{i, j}$, is assumed to follow a $Beta(\mu_{i, j_{k}}, \phi_{i, j_{k}})$ distribution, where $\mu_{i, j_{k}}$ is the mean and $\phi_{i, j_{k}}$ is the dispersion of the beta distribution[^beta_parameterisation]. Both the means, $\mu_{i, j_{k}}$, and the dispersions, $\phi_{i, j_{k}}$, are group-specific and are allowed to vary across methylation loci ($i$). The dispersion parameter models the within-group variability of the $B_{i, j}$, i.e. the within-group _biological variability_ of DNA methylation.

@@

As noted by \citet{Feng:2014iq}, the beta distribution is a very flexible distribution with support on $[0, 1]$ and has "long been a natural choice to model binomial proportions", particularly as a conjugate prior, as it is used in the empirical Bayes model of \citet{Feng:2014iq}. The beta-binomial model can also be viewed from a non-Bayesian perspective as a compound distribution or overdispersed binomial distributions

@@

An alternative to specifying a parametric distribution for the $B_{i, j}$ is to sample these from real data, e.g., by sampling some observed $\beta_{i, j}$ for a particular dataset and treating them as if they were observations on $B_{i, j}$. `WGBSSuite`\citep{Rackham:2015bv} uses a modified form of this approach. In `WGBSSuite`, a hidden Markov model is used to classify every CpG as having an underlying state ("de-methylated", "1st transition", "2nd transition" or "methylated"). Each of these four states has a region-specific average methylation level that is based on the distribution of $\beta$-values for a chosen dataset. For example, the average methylation level for all "de-methylated" regions is defined as $B_{de-methylated} = median(\{ \beta_{i, j}: \beta_{i, j} \in [0, 0.5) \}])$. Then, each $B_{i, j}$ is a perturbed version of this region-level average methylation, $B_{region-type}$, obtained by adding on a zero-mean Gaussian random variable, i.e. $B_{i, j} = B_{region-type} + \epsilon_{i, j}$, where $\epsilon_{i, j} \eqd Normal(0, s^{2})$. Care needs to be taken that $0 \leq B_{i, j} \leq 1$.

@@

The sequencing depth at each methylation loci, $d_{i, j}$, may be sampled from real data \citep[e.g.]{Feng:2014iq, Chen:2014jb, Dolzhenko:2014bo}, or simulated from a parametric distribution such as the Poisson \citep{Rackham:2015bv}, a rounded Gaussian distribution \citep{Xu:2013eg}, or a rounded mixture of Gamma distributions \citep{Lacey:2013iy}.

The most sophisticated approach to simulation of sequencing depth in bisulfite-sequencing experiments is given by \citet{Lacey:2013iy}. In addition to using a mixture of distributions to capture both the low-coverage and high-coverage modes observed in RRBS sequencing coverage, \citet{Lacey:2013iy} model the correlation of sequencing depth across samples for a given region. They do this by using a Normal copula to make the set of sequencing depths a jointly dependent set of random variables. While this is undoubtably sophisticated, the effect of correlated versus uncorrelated sequencing depths in a simulation model is not explored in the paper and so the cost-benefit trade-off is unclear. Moreover, it is simpler, and likely more computationally efficient, to include such correlations by a sensible sampling of sequencing depths from real data.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/dmr_break.pdf}
\caption[A hypothetical differentially methylated region]{A hypothetical region in a study of differential methylation in a two-group experiment. Plotted is the difference in $\beta$-value between the two groups ($\Delta \beta$) with associated standard error against the position along the genome (\emph{Position (bp)}). The first three methylation loci are DMLs, as are the last five methylation loci. However, the fourth locus is not a DML. Should this region be considered as two distinct DMRs or as a single DMR?}
\label{fig:dmr_break}
\end{figure}

@@

Most simulation methods do not simulate individual reads and so cannot simulate within-fragment co-methylation. Instead, they capture correlations of methylation levels by inducing dependence in the $B_{i, j}$. For example, under the Beta-Binomial model these correlations could be induced by forcing the $\mu_{i, j}$ to be spatially dependent. This idea takes its inspiration from \citet{Jaffe:2012gx} who simulate spatially correlated DNA methylation microarray data by imposing an autocorrelation structure via a lag-1 autoregressive model of the simulated $\beta$-values.

@@

`WGBSSuite` induces spatial correlation amongst the $B_{i, j_{k}}$ in a less direct manner. Recall that each methylation locus is assigned one of the four underlying states ("de-methylated", "1st transition", "2nd transition", "methylated") via a hidden Markov model. The transition matrix of this hidden Markov model ensures that neighbouring loci, $i, (i+1)$, are more likely to be assigned the same state. Furthermore, since all loci within each of the four states are assigned the same (perturbed) underlying methylation level, $B_{i, j_{k}} = B_{region-type} + \epsilon_{i, j_{k}}$, neighbouring loci have similar methylation levels. Note that the $IPD$ only plays a direct role in the initial segmentation of the genome, not in the assigning of the $B_{i, j_{k}}$.

@@

A separate class of model-based simulation methods are those that simulate the $\beta$-values directly, i.e. without simulating sequencing depth \citet[e.g.,][]{Jaffe:2012gx, Chen:2013eh, Chen:2014jb}. These models are designed for simulating microarray data and not sequencing data, since they do not include the variability due to variation in sequencing depth. Of these methods, only \citet{Jaffe:2012gx} simulate correlation amongst the $B_{i, j_{k}}$.

@@

A simulation may also be based entirely on re-sampling of real data[^synthetic]. This type of simulation is attractive because through careful sampling it can capture behaviour that is otherwise very difficult, if not impossible, to capture in a parametric model. At the same time, however, if the sampling units are poorly chosen or the sampling strategy is incorrect, then it may ignore these same features or, worse still, introduce artefacts into the simulated data.

@@

Re-sampling methods are most easily implemented at the level of $\beta$-values. Re-sampling reads is reads is more difficult, except in the special case of down-sampling whereby the positions of reads are held constant but only a sub-sample of them are used in downstream analysis. Down-sampling is only really of interest to examine the effects of sequencing coverage on downstream analyses.

@@

Due to the correlation structure of the $\beta$-values, the co-methylation, it is likely, although not guaranteed, that other CpGs in the blocks containing targets also display differential methylation.

@@

`methsim` is specifically designed to model the co-methylation structure of bisulfite-sequencing data by simulating individual DNA fragments rather than directly simulating summary methylation measurements. In order to model the co-methylation structure, I decided that `methsim` should simulate at the level of individual DNA fragments. An added bonus of this approach is that `methsim` can (in theory) generate data at multiple resolutions: $\beta$-values, methylation patterns at m-tuples or entire sequencing reads. While my initial motivation in developing `methsim` was to explore models of co-methylation, I realised that it could also be used in the development and comparison of methods for the downstream analysis of DNA methylation data.

@@

1. Simulating the true methylome of each sample.
2. Simulating reads, including sequencing error and bisulfite-conversion error, by sampling from the true methylome.
3. Constructing the output, be it reads, methylation patterns at m-tuples or $\beta$-values.

@@

`methsim` makes extensive use of the S4 object system in R. The most important classes defined by `methsim` are `MethylomeParam`, `SimulatedMethylome`, `WGBSParam` and `SimulatedBS`. The most important methods are the `simulate()` methods defined for the `MethylomeParam` and `WGBSParam` classes.

The `MethylomeParam` object contains the empirical distributions of key statistics from which the parameters for simulating the 'true' methylome are sampled. Therefore, the `MethylomeParam` should be based on data from a relevant sample. To help a new user get started, `methsim` includes `MethylomeParam` objects for the _ADS_, _ADS-adipose_ and _ADS-iPSC_ samples from the _Lister_ dataset. Alternatively, the user may process the `BAM` file for their sample with `methtuple` and then use the helper functions in `methsim` to construct a `MethylomeParam` object based on their sample of interest.

To simulate a true methylome, the user runs the `simulate()` method on the `MethylomeParam` object. This returns a `SimulatedMethylome` object.

@@

To simulate values of $\mathbf{B}$ and $\bm{\psi}$, `methsim` uses the empirical distributions of their estimates, $\bm{\beta}$ and $\widehat{\bm{\theta}} = \log(\widehat{\bm{\psi}})$, in the input methylome. More specifically, $\mathbf{B}$ is based on $\beta_{i | r_{i}}$ (the empirical distribution of $\beta_{i}$ conditional on the region type of the $i^{th}$ locus, $r_{i}$) and $\bm{\theta}$ is based on $\widehat{\theta}_{MH}$. We have already seen many examples of the distribution of $\widehat{\theta}_{MH}$ in Chapter \ref{chap:co-methylation}. The empirical distributions $\beta_{i | r_{i}}$ for the _ADS_ sample are shown in Figure \ref{fig:ADS_beta_by_pm_boxplots}.

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/ADS_beta_by_pm_boxplots.pdf}
\caption[Distribution of $\beta$-values stratified by region type]{$\beta_{i | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the \emph{ADS} sample. Only CpGs with at least $10\times$ sequencing coverage are used. UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
\label{fig:ADS_beta_by_pm_boxplots}
\end{figure}

@@

\begin{description}
  \item[\emph{m1}] All loci in each region have the same $B$, which is sampled from $\beta_{i | r_{i}}$.
  \item[\emph{m2}] All loci in each region have the same \textbf{average} $B$, which is sampled from $\beta_{i | r_{i}}$; this average is then perturbed for each locus by a $Gaussian(0, \sigma_{B}^2)$ random variable. This is similar to what is done by \texttt{WGBSsuite}. Here I have used $\sigma_{B}^{2} = B \times (1 - B)$. The resulting $\mathbf{B}$ is truncated so that all values lie between 0.01 and 0.99 (necessary to avoid issues in the iterative proportional fitting algorithm at the boundaries).
  \item[\emph{m3}] $\mathbf{B}$ simulated as in \emph{m2} but with the perturbations in each region simulated from a first-order autoregressive process with coefficient equal to $0.5$.
  \item[\emph{A}] $\bm{\theta}$ identically zero. This is to simulate independence of within-fragment methylation states and as a control to check that \texttt{methsim} is working as intended for simulating within-fragment co-methylation.
  \item[\emph{B}] $\theta$ sampled from $Gaussian(\mu_{IPD}, \sigma_{IPD}^{2})$, where $\mu_{IPD}$ and $\sigma_{IPD}$ are plug-in estimates computed from the distribution of chromosome-level $\widehat{\theta}_{MH}$ (autosomes only).
\end{description}

@@

`methsim` uses a simple Poisson-based method for simulating bisulfite-sequencing. The user specifies the desired read length[^read_length], the average sequencing coverage, the error rate ($\epsilon$, which includes both sequencing error and bisulfite-conversion error), and provides the simulated 'true' methylome. We will use $200$ bp, single-end reads (equivalently, $100$ bp paired-end reads that are always end-to-end) with an average coverage of $30 \times$ in what follows.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methsim_plotBetaFreqPoly_CGI.pdf}
\caption[\texttt{methsim}: $\beta$-value kernel density stratified by CpG islands]{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \texttt{methsim} simulated data and the \emph{ADS} input sample, stratified by whether the CpG is in a CpG island. Only CpGs with at least $10 \times$ sequencing coverage are included. `Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
\label{fig:methsim_plotBetaFreqPoly_CGI}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methsim_MH_by_seqnames.pdf}
\caption[\texttt{methsim}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \texttt{methsim} simulated data and the \emph{ADS} input sample using $NIL = 0$ pairs. The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.}
\label{fig:methsim_MH_by_seqnames}
\end{figure}

>>>>

There is a lack of gold-standard datasets in the field of DNA methylation for the benchmarking and validation of analysis methods. Therefore, the development of realistic, cheap, and user-friendly software to simulate DNA methylation data will be of benefit.

@@

`Sherman` is software to simulate bisulfite-sequencing reads as `FASTQ` files, including various 'contaminants', such as SNPs, basecall errors and sequence artefacts (\url{http://www.bioinformatics.babraham.ac.uk/projects/sherman/}). The simulated reads are designed for comparing the performance of different alignment strategies. `Sherman` has many parameters, of which the ones relevant to our discussion of simulating realistic DNA methylation data are `-CG` and `-CH`, the bisulfite conversion rates for CG and CH methylation loci, respectively. These are set by the user with values between $0$ and $100$ ($\%$). Reads are simulated by sampling from the user-specified reference genome. When a read contains a CG (resp. CH) locus, it is randomly assigned as being converted to a TG (resp. TH) with probability `-CG` / 100 (resp. `-CH` / 100).

@@

The most widely studied downstream analysis problem is that of identifying differential methylation, which is done by comparing summary measures of methylation, such as $\beta$-values, between two or more groups. It is therefore generally sufficient to directly simulate these aggregated measures, rather than simulating reads, for these type of simulation studes.

Most papers that propose a new method for downstream analysis of bisulfite-sequencing data include a simulation study. Generally, such a simulation method exists to support claims about the performance of the proposed method and is not a major feature of the paper. Consequently, the simulation model is often only briefly described and a software implementation is rarely made available. In fact, until recently, of the methods reviewed in this section, only that of \citet{Lacey:2013iy} had a software implementation available. As I was writing this chapter, the `WGBSSuite` software was published \citep{Rackham:2015bv}. `WGBSSuite` is the only software specifically published for the purpose of simulating whole-genome bisulfite-sequencing data for comparing methods for identifying differential methylation. `WGBSSuite` is available for download as a collection of R scripts[^WGBSSuite].

@@

Almost all downstream analyses are reference-based (see Section \ref{sec:methylated_or_unmethylated}), so it is desirable to know how these methods perform with respect to the relevant reference genome, not a simulated genome in which the location of methylation loci vary from simulation to simulation.

@@

Recall that $B_{i, j}$ is the underlying 'true' methylation level at locus $i$ in sample $j$. In the context of simulating data for comparing differential methylation calling methods, we want these to be group-specific. That is, we want to specify $B_{i, j_{k}}$, where $j_{k}$ indicates that sample $j$ comes from group $k$. For non-differentially methylated loci all $k$ groups have identical $B_{i, j_{k}} = B_{i, j_{0}}$; differentially methylated loci are simulated by setting $B_{i, j_{k}} \neq B_{i, j_{k'}}$ for some $k \neq k'$.

Under the beta-binomial model, the true methylation level, $B_{i, j}$, is assumed to follow a $Beta(\mu_{i, j_{k}}, \phi_{i, j_{k}})$ distribution, where $\mu_{i, j_{k}}$ is the mean and $\phi_{i, j_{k}}$ is the dispersion of the beta distribution[^beta_parameterisation]. Both the means, $\mu_{i, j_{k}}$, and the dispersions, $\phi_{i, j_{k}}$, are group-specific and are allowed to vary across methylation loci (i.e. across $i$). The dispersion parameter models the within-group variability of the $B_{i, j}$, i.e. the within-group _biological variability_ of DNA methylation.

@@

As noted by \citet{Feng:2014iq}, the beta distribution is a very flexible distribution with support on $[0, 1]$ and has "long been a natural choice to model binomial proportions", particularly as a conjugate prior, as it is used in the empirical Bayes model of \citet{Feng:2014iq}. The beta-binomial model can also be viewed from a non-Bayesian perspective as a compound distribution or as an overdispersed binomial distribution.

@@

An alternative to specifying a parametric distribution for the $B_{i, j}$ is to sample these from real data, e.g., by sampling some observed $\beta_{i, j}$ for a particular dataset and treating them as if they were observations on $B_{i, j}$. `WGBSSuite` \citep{Rackham:2015bv} uses a modified form of this approach. In `WGBSSuite`, a hidden Markov model is used to classify every CpG as having an underlying state ("de-methylated", "1st transition", "2nd transition" or "methylated"). Each of these four states has a region-specific average methylation level that is based on the distribution of $\beta$-values for a chosen dataset. For example, the average methylation level for all "de-methylated" regions is defined as $B_{de-methylated} = median(\{ \beta_{i, j}: \beta_{i, j} \in [0, 0.5) \}])$. Then, each $B_{i, j}$ is a perturbed version of this region-level average methylation, $B_{region-type}$, obtained by adding on a zero-mean Gaussian random variable, i.e. $B_{i, j} = B_{region-type} + \epsilon_{i, j}$, where $\epsilon_{i, j} \eqd Normal(0, s^{2})$. Care needs to be taken that $0 \leq B_{i, j} \leq 1$.

@@

The sequencing depth at each methylation loci, $d_{i, j}$, may be sampled from real data \citep{Feng:2014iq, Chen:2014jb, Dolzhenko:2014bo}, or simulated from a parametric distribution such as the Poisson \citep{Rackham:2015bv}, a rounded Gaussian distribution \citep{Xu:2013eg}, or a rounded mixture of Gamma distributions \citep{Lacey:2013iy}.

The most sophisticated approach to simulation of sequencing depth in bisulfite-sequencing experiments is given by \citet{Lacey:2013iy}. In addition to using a mixture of distributions to capture both the low-coverage and high-coverage modes observed in RRBS sequencing coverage, \citet{Lacey:2013iy} model the correlation of sequencing depth across samples for a given region. They do this by using a Gaussian copula to make the set of sequencing depths a jointly dependent set of random variables. While this is undoubtably sophisticated, the effect of correlated versus uncorrelated sequencing depths in a simulation model is not explored in the paper and so the cost-benefit trade-off is unclear. Moreover, it is simpler, and likely more computationally efficient, to include such correlations by a sensible sampling of sequencing depths from real data.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/dmr_break.pdf}
\caption[A hypothetical differentially methylated region]{A hypothetical region in a study of differential methylation in a two-group experiment. Plotted is the difference in $\beta$-value between the two groups ($\Delta \beta$) with associated standard error against the position along the genome (\emph{Position (bp)}). The first three methylation loci are DMCs, as are the last five methylation loci. However, the fourth locus is not a DMC. Should this region be considered as two distinct DMRs or as a single DMR?}
\label{fig:dmr_break}
\end{figure}

@@

Most simulation methods do not simulate individual reads and so cannot simulate within-fragment co-methylation. Instead, they capture correlations of methylation levels by inducing dependence in the $B_{i, j}$. For example, under the beta-binomial model these correlations could be induced by forcing the $\mu_{i, j}$ to be spatially dependent. This idea takes its inspiration from \citet{Jaffe:2012gx} who simulate spatially correlated DNA methylation microarray data by imposing an autocorrelation structure via a lag-1 autoregressive model of the simulated $\beta$-values.

@@

`WGBSSuite` induces spatial correlation amongst the $B_{i, j_{k}}$ in a less direct manner. Recall that each methylation locus is assigned one of the four underlying states ("de-methylated", "1st transition", "2nd transition", "methylated") via a hidden Markov model. The transition matrix of this hidden Markov model ensures that neighbouring loci, $i, (i+1)$, are more likely to be assigned the same state. Furthermore, since all loci within each of the four states are assigned the same (perturbed) underlying methylation level, $B_{i, j_{k}} = B_{region-type} + \epsilon_{i, j_{k}}$, neighbouring loci have similar methylation levels. Note that the $IPD$ only plays a direct role in the initial segmentation of the genome, not in the assigning of the $B_{i, j}$.

@@

A separate class of model-based simulation methods are those that simulate the $\beta$-values directly, i.e. without simulating sequencing depth \citep{Jaffe:2012gx, Chen:2013eh, Chen:2014jb}. These models are designed for simulating microarray data and not sequencing data, since they do not include the variability due to variation in sequencing depth. Of these methods, only \citet{Jaffe:2012gx} simulate correlation amongst the $B_{i, j_{k}}$.

@@

A simulation may also be based entirely on re-sampling of real data[^synthetic]. This type of simulation is attractive because, through careful sampling, it can capture behaviour that is otherwise very difficult, if not impossible, to capture in a parametric model. At the same time, however, if the sampling units are poorly chosen or the sampling strategy is incorrect, then it may ignore these same features or, worse still, introduce artefacts into the simulated data.

@@

Re-sampling methods are most easily implemented at the level of $\beta$-values. Re-sampling reads is more difficult, except in the special case of down-sampling whereby the positions of reads are held constant but only a sub-sample of them are used in downstream analysis. Down-sampling is only really of interest to examine the effects of sequencing coverage on downstream analyses.

@@

Due to the correlation structure of the $\beta$-values (the co-methylation) it is likely, although not guaranteed, that other CpGs in the blocks containing targets also display differential methylation.

@@

`methsim` is specifically designed to model the co-methylation structure of bisulfite-sequencing data by simulating individual DNA fragments rather than directly simulating summary methylation measurements. In order to model the co-methylation structure, I decided that `methsim` should simulate at the level of individual DNA fragments. An added bonus of this approach is that `methsim` can (in theory) generate data at multiple resolutions: $\beta$-values, methylation patterns at m-tuples, or entire sequencing reads. While my initial motivation in developing `methsim` was to explore models of co-methylation, I realised that it could also be used in the development and comparison of methods for the downstream analysis of DNA methylation data.

@@

1. Simulating the true methylome of each sample.
2. Simulating reads, including sequencing error and bisulfite-conversion error, by sampling from the true methylome.
3. Constructing the output, be it reads, methylation patterns at m-tuples, or $\beta$-values.

@@

`methsim` makes extensive use of the S4 object system in R. The most important classes defined in `methsim` are `MethylomeParam`, `SimulatedMethylome`, `WGBSParam` and `SimulatedBS`. The most important methods are the `simulate()` methods defined for the `MethylomeParam` and `WGBSParam` classes.

The `MethylomeParam` object contains the empirical distributions of key statistics from which the parameters for simulating the 'true' methylome are sampled. Therefore, the `MethylomeParam` object should be based on data from a relevant sample. To help a new user get started, `methsim` includes `MethylomeParam` objects for the _ADS_, _ADS-adipose_ and _ADS-iPSC_ samples from the _Lister_ dataset. Alternatively, the user may process the `BAM` file for their sample with `methtuple` and then use the helper functions in `methsim` to construct a `MethylomeParam` object based on their sample of interest. The user runs the `simulate()` method on the `MethylomeParam` object to simulate a true methylome; this returns a `SimulatedMethylome` object.

@@

To simulate values of $\mathbf{B}$ and $\bm{\psi}$, `methsim` uses the empirical distributions of their estimates, $\bm{\beta}$ and $\widehat{\bm{\theta}} = \log(\widehat{\bm{\psi}})$, in the input methylome. More specifically, $\mathbf{B}$ is based on $\bm{\beta_{i | r_{i}}}$ (the empirical distribution of $\beta_{i}$ conditional on the region type of the $i^{th}$ locus, $r_{i}$) and $\bm{\theta}$ is based on $\widehat{\theta}_{MH}$. We have already seen many examples of the distribution of $\widehat{\theta}_{MH}$ in Chapter \ref{chap:co-methylation}. The empirical distributions $\bm{\beta_{i | r_{i}}}$ for the _ADS_ sample are shown in Figure \ref{fig:ADS_beta_by_pm_boxplots}.

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/ADS_beta_by_pm_boxplots.pdf}
\caption[Distribution of $\beta$-values stratified by region type]{$\bm{\beta_{i | r_{i}}}$, the distribution of $\beta$-values for CpGs in each region type, for the \emph{ADS} sample. Only CpGs with at least $10\times$ sequencing coverage are used. UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
\label{fig:ADS_beta_by_pm_boxplots}
\end{figure}

@@

\begin{description}
  \item[\emph{m1}] All loci in each region have the same $B$, which is sampled from $\bm{\beta_{i | r_{i}}}$.
  \item[\emph{m2}] All loci in each region have the same \textbf{average} $B$, which is sampled from $\bm{\beta_{i | r_{i}}}$; this average is then perturbed for each locus by a $Gaussian(0, \sigma_{B}^2)$ random variable. This is similar to what is done by \texttt{WGBSsuite}. Here I have used $\sigma_{B}^{2} = B \times (1 - B)$. The resulting $\mathbf{B}$ is truncated so that all values lie between 0.01 and 0.99 (necessary to avoid issues in the iterative proportional fitting algorithm at the boundaries).
  \item[\emph{m3}] $\mathbf{B}$ simulated as in \emph{m2} but with the perturbations in each region simulated from a first-order autoregressive process with coefficient equal to $0.5$.
  \item[\emph{A}] $\bm{\theta}$ identically zero. This is to simulate independence of within-fragment methylation states and as a control to check that \texttt{methsim} is working as intended for simulating within-fragment co-methylation.
  \item[\emph{B}] $\theta$ sampled from $Gaussian(\mu_{IPD}, \sigma_{IPD}^{2})$, where $\mu_{IPD}$ and $\sigma_{IPD}$ are plug-in estimates computed from the distribution of chromosome-level $\widehat{\theta}_{MH}$ (autosomes only).
\end{description}

@@

`methsim` uses a simple Poisson-based method for simulating bisulfite-sequencing. The user specifies the desired read length[^read_length], the average sequencing coverage, the error rate ($\epsilon$, which includes both sequencing error and bisulfite-conversion error), and provides the simulated 'true' methylome. We will use $200$ bp, single-end reads (equivalently, $100$ bp paired-end reads that are always end-to-end) with an average coverage of $23 \times$ (mimicking the actual _ADS_ data) in what follows.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methsim_plotBetaFreqPoly_CGI.pdf}
\caption[\texttt{methsim}: $\beta$-value frequency polygon stratified by CpG islands]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \texttt{methsim} simulated data and the \emph{ADS} input sample, stratified by whether the CpG is in a CpG island. Only CpGs with at least $10 \times$ sequencing coverage are included. `Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
\label{fig:methsim_plotBetaFreqPoly_CGI}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methsim_MH_by_seqnames.pdf}
\caption[\texttt{methsim}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \texttt{methsim} simulated data and the \emph{ADS} input sample using $NIL = 0$ pairs. The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points (only autosomal data are simulated, hence only autosomal data are shown).}
\label{fig:methsim_MH_by_seqnames}
\end{figure}
